\bf Baseline Drift Estimation for Time Series Data Using Quantile Trend Filtering

### 2019-02-05 Manuscript/detrendify_JCGS.tex
incorporated Joe's comments on sections 1 and 2

	\if0\blind
	{
		\title{\bf Baseline Drift Estimation for Time Series Data Using Quantile Trend Filtering}
		\author{Halley Brantley\thanks{
				Department of Statistics, North Carolina State University, Raleigh, NC 27695 (E-mail: hlbrantl@ncsu.edu)} \,
			Joseph Guinness\thanks{
				Department of Biological Statistics and Computational Biology, Cornell University, Ithaca, NY 14853 (E-mail: guinness@cornell.edu )} \,
			and
			%    and
			Eric C. Chi\thanks{Department of Statistics, North Carolina State University, Raleigh, NC 27695 (E-mail: eric$\_$chi@ncsu.edu).}    \\}
		\date{}
		\maketitle
	} \fi

@@

	\bigskip
	\begin{abstract}
		We address the problem of estimating smoothly varying baseline trends in time series data using quantile trend filtering. We extend the basic framework to ensure non-crossing while estimating multiple trends simultaneously and implement an alternating direction method of moments (ADMM) algorithm for series that cannot be processed simultaneously. Timing experiments demonstrate the usefulness of our ADMM algorithm even in cases where the series can be processed simultaneously. We also address smoothing parameter selection and propose a modified criteria based on the extended Bayesian Information Criteria. We demonstrate both through simulation studies and an application to low cost air quality sensor data that our model provides better or comparable estimates than existing methods of non-parametric quantile trends and improves signal classification of low-cost air quality sensor output. 
	\end{abstract}

@@

	In applications spanning the fields of chemistry \citep{Ning2014}, macroeconomics \citep{yamada2017estimating}, environmental science \citep{brantley2014mobile}, and medical sciences \citep{pettersson2013algorithm, marandi2015qualitative}, scalar time series are observed and assumed to consist of a slowly varying trend and other more rapidly varying components. \cite{Kim2009} addressed the case when the series consisted of just a trend and a random noise component and proposed the use of \textit{$\ell_1$ trend filtering} to estimate trends that piecewise linear or piecewise polynomial. \cite{Tib2014} later demonstrated that empirically the trend filtering estimates adapt to the local level of smoothness better than the more common smoothing splines. In the trend filtering problem, given observations $y(t)$ with $t=1,...,N$, the trend, $\theta(t)$, is estimated by solving the following convex problem.
	\begin{eqnarray}
	\underset{\theta}{\arg\min}\; \frac{1}{2} \lVert y - \theta \rVert_2^2 + \lambda \lVert \mathbf{D}^{(k+1)}\theta \rVert_1,
	\end{eqnarray}
	where $\lambda \geq 0$ is a regularization parameter and the matrix $\mathbf{D}^{(k+1)} \in \mathbb{R}^{(n - k -1) \times n}$ is the discrete difference operator of order $k+1$. To understand the purpose of penalizing $\mathbf{D}^{(k+1)}$ consider the difference operator when $k = 0$.
	\begin{eqnarray}
	\mathbf{D}^{(1)} = \begin{pmatrix}
	-1 & 1 & 0 & \cdots & 0 & 0 \\
	0 & -1 & 1 & \cdots & 0 & 0 \\
	\vdots & & & & & \\
	0 & 0 & 0 & \cdots & -1 & 1 \\
	\end{pmatrix}
	\end{eqnarray}
	Thus, $\lVert \mathbf{D}^{(1)}\theta \rVert_1 = \sum_{i=1}^{n-1} \lvert \theta_i - \theta_{i+1} \rvert$ which is just total variation denoising in one dimension. The penalty incentivizes solutions which are piece-wise constant. For $k \geq 1$, the difference operator $\mathbf{D}^{(k+1)} \in \mathbb{R}^{(n-k-1) \times n}$ is defined recursively as follows
	\begin{eqnarray}
	\mathbf{D}^{(k+1)} & = & \mathbf{D}^{(1)}\mathbf{D}^{(k)}.
	\end{eqnarray}
	By penalizing the $k+1$ fold composition of the discrete difference operator, we obtain solutions which are piecewise polynomials of order $k$.

	Trend filtering provides excellent estimates of trends when the only components of the time series are the trend and random noise (Fig. \ref{fig:trendfilter}). In some applications, however, time series may also include a rapidly varying signal component in addition to the trend and noise, in these cases the trend filtering estimate over-estimates the trend in the places where signal is present (Fig. \ref{fig:trendfilter}).

	\begin{figure}
		\centering
		\caption{Examples of trend filtering solutions. The true trend is shown in black, while the estimated trend is shown in red.}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg1.png}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg2.png}
		\label{fig:trendfilter}
	\end{figure}

	One application in which the observed time series consists of a slowly varying baseline, non-negative signal, and rapidly varying noise is the output of low cost air quality sensors. The use of low-cost, portable, air quality sensors has increased dramatically in the last decade. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. To reduce cost and power consumption of the SPod, the temperature and relative humidity of the air presented to the photoionization detectors (PIDs) is not controlled and as a result the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Fig. \ref{fig:raw_spod} provides an example of 3 co-located SPod sensor outputs from near an industrial facility. While all of the sensors response to the pollutant signal, the amount of baseline drift varies from one node to another.

	\begin{figure}[b!]
		\caption{Example of 3 co-located SPod PID sensor readings.}
		\includegraphics[width = \linewidth]{Figures/uncorrected_data.png}
		\label{fig:raw_spod}
	\end{figure}

	In this application, as well as those described in \cite{Ning2014}, \cite{marandi2015qualitative}, and \cite{pettersson2013algorithm}, it is not the mean trend that is desired but rather the baseline trend, which can be thought of as a trend in a low quantile of the data.  \cite{Koenker1978} were the first to propose substituting the squared loss typically used in regression with the check loss function  (Eq. \ref{eq:checkloss}) to estimate a conditional quantile instead of the conditional mean.

	\begin{equation}
	\label{eq:checkloss}
	 \rho_{\tau}(z) = \sum_{i=1}^n z_i(1-\mathbf{I}(z_i<0)).
	\end{equation}

	Later, \cite{KoenkerNgPortnoy1994} addressed the issue of estimated quantile curves by combining the check loss penalty with trend filtering penalty with $k = 2$ which produces quantile trends that are piecewise linear. Other approaches to estimating quantile trends have also been proposed. Rather than using the trend filtering $\ell_1$-norm to penalize the discrete differences, \cite{nychka1995nonparametric} used the smoothing spline penalty based on the $\ell_2$-norm solving the optimization problem in Eq. \ref{eq:smoothingspline}.

	\begin{equation*}
	\label{eq:smoothingspline}
	\rho_{\tau}(y - f) + \lambda\int (f''(t))^2 dt, 
 	\end{equation*}

	where $f(t)$ is a smooth function of time and $\lambda$ is a tuning parameter that controls the degrees of smoothing. \cite{Oh2011} proposed an algorithm for solving this problem based on approximating the checkloss function with a similar differentiable function.

	\cite{Racine2017} take an entirely different approach to estimating quantile trends. They constrain the response to follow a smooth location scale model of the form $y(t) = a(t) + b(t)\epsilon_i$ and estimate the $\tau$\textsubscript{th} conditional quantile given $t = t_0$ using a kernel estimator and local linear approach.

	We propose to use the trend filtering penalty combined with the check loss function to produce quantile trends that are piecewise quadratic. The formulation was proposed by \cite{Kim2009} as a possible extension of $\ell_1$-trend filtering but not studied. Moreover we extend the basic framework to ensure non-crossing while modeling multiple quantiles. We also implement a parallel ADMM algorithm for series that are too large to be computed simultaneously and propose a modified criteria for choosing the smoothing parameter. We demonstrate through simulation studies that our proposed model provides better or comparable estimates of non-parametric quantile trends than existing methods and is a more effective method of drift removal for low-cost air quality sensors.

	\section{Methods}

	\subsection{Quantile Trend Filtering}

	We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$ the estimation of the quantile trend filtering model can be posed as the following optimization problem.
	\begin{eqnarray}
	\label{eq:quantile_trend}
	\underset{\theta}{\min}\; \rho_\tau(y - \theta) + \lambda \lVert \mathbf{D}^{(k+1)} \theta \rVert_1,
	\end{eqnarray}
	where $\lambda$ is a non-negative regularization parameter. We address the problem of choosing $\lambda$ in Section \ref{sec:lambda_choice}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of free or commercial solvers. In many cases, including ours, we are interested in estimating multiple quantiles simultaneously. We also want to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$. Given quantiles $\{\tau_1, ..., \tau_J\}$ such that $\tau_1 < \tau_2 < ... < \tau_J$, the optimization problem becomes

	\begin{eqnarray}
	\label{eq:noncross_trend}
	\underset{\theta_1, ..., \theta_J}{\min}\; \sum_{j=1}^J \left [\rho_{\tau_j}(y - \theta_{j}) + 
	\lambda_j \lVert \mathbf{D}^{(k+1)} \theta_j \rVert_1 \right ] \\
	 \text{subject to: }\; \theta_{1}(t) \le \theta_{2}(t) \le ... \le \theta_{J}(t) \text{ for all } t,
	\end{eqnarray}

	where $\theta_j \in \mathcal{R}^n$. The additional constraints are linear in the parameters so the non-crossing quantile trends can still be estimated by a number of available solvers. In the rest of this paper we rely on the commercial solver Gurobi \citep{gurobi} and its R package implementation, but it can easily be substituted for a free solver such as the Rglpk package by \cite{rglpk}. The number of parameters to be estimated in this problem is equal to the number of observations multiplied by the number of quantiles of interest. As the size of the data and the number of quantiles grows, all solvers will eventually break.

@@

	To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose an alternating direction method of multipliers (ADMM) algorithm for solving large problems in a piecewise fashion. The ADMM algorithm \citep{gabay1975dual, glowinski1975approximation} is fully described by \cite{boyd2011distributed}. We apply the consensus ADMM algorithm to the the quantile regression trend filtering problem given in Eq. \ref{eq:quantile_trend}, by dividing our observed series $y(t)$ with $t = \{1, ..., n\}$ into overlapping windows

	\begin{align*}
	y_1(t) = y(t) & \mbox{~~if~~} 1 \le t \le u_{1}\\
	y_2(t) = y(t) & \mbox{~~if~~} l_{2} \le t \le u_{2} \\
	\cdots & \\
	y_M(t) = y(t) & \mbox{~~if~~} l_{M} \le t \le  n\\
	\end{align*}

	 with boundaries $1 < l_{2} < u_{1} < l_{3} < u_{2} < l_{4} < u_{3} < ...< n$. An illustration is given in Fig. \ref{fig:windows}. Given quantiles $\tau_1 < ... < \tau_J$ to be estimated we define $\theta_{j,m}(t)$ as the value of the $\tau_j$\textsuperscript{th} quantile trend in window $M$ at time point $t$. In order to write out the constraint that the overlapping sections must be equal we define a consensus variable

	 \begin{align*}
		 \bar{\theta}_{j,m} =&  g(\theta_{j, m-1}, \theta_{j,m}, \theta_{j,m+1}) \\
		 =& \begin{cases} 
			 \frac{\theta_{j,m-1}(t)+\theta_{j,m}(t)}{2} & \mbox{if~~} l_{m} \le t \le u_{m-1}  \\
			 \theta_{j,m}(t) & \mbox{if~~} u_{m-1} \le t \le l_{m+1}  \\
			 \frac{\theta_{j,m}(t)+\theta_{j,m+1}(t)}{2} & \mbox{if~~} l_{m+1} \le t \le u_{m}  
			 \end{cases},
	\end{align*}
	defining $\theta_{j,M+1} = \theta_{j,M}$ and $\theta_{j,0} = \theta_{j,1}$. Our windowed quantile trend optimization problem can then be written as 
	 \begin{align*}
		 \label{eq:quantile_windows}
		 &\sum_{m=1}^M\underset{\theta_{1,m}, ..., \theta_{J,m}}{\min}\; \sum_{j=1}^J \left [\rho_{\tau_j}(y_m - \theta_{j,m}) + 
		 \lambda_j \lVert \mathbf{D}^{(k+1)} \theta_{j,m} \rVert_1 \right ] \\
		 &\text{subject to: }\; \\
		 &\qquad \theta_{1,m}(t) \le \theta_{2,m}(t) \le ... \le \theta_{J,m}(t) \text{ for all } m,t \\
		 &\qquad \theta_{j,m}(t) = \bar{\theta}_{j,m}(t) \text{ for all } j, m, t
	 \end{align*}
	 Defining Lagrange multipliers $\omega_{j,m}$, and penalty parameter $\gamma > 0$ we can write the augmented Lagrangian for finding the trends in window $m$:
	 \begin{equation*}
	 \mathcal{L}(\theta_{j,m}, \bar{\theta}_{j,m}, \omega_{j,m}) = \sum_{j=1}^J\rho_{\tau_j}(y_m - \theta_{j,m})+\lambda \lVert \mathbf{D}^{(k+1)}\theta_{j,m}\rVert_1 +  \omega_{j,m}^T(\theta_{j,m} - \bar{\theta}_{j,m}) + 
	 \frac{\gamma}{2}||\theta_{j,m} - \bar{\theta}_{j,m}||_2^2
	 \end{equation*}
	 We then estimate the trend separately in each window, which can be done in parallel, while constraining the overlapping pieces of the trends to be equal as outlined in algorithm 1.

	\begin{figure}[!h] 
		\centering
		\caption{Window boundaries and trends fit separately in each window.}
		\includegraphics[width = 0.8\linewidth]{Figures/overlapping_windows.png}
		\label{fig:windows}
	\end{figure}

@@

	We measure convergence use the stopping criteria described by \cite{boyd2011distributed}. The criteria are based on the primal and dual residuals which represent the residuals for the primal and dual feasibility, respectively. The primal residual, 
	\begin{equation}
	r_p^{(q)} = \sqrt{\sum_{m=1}^M\sum_{j=1}^J\lVert\theta_{j,m}^{(q)} - \bar{\theta}_{j,m}^{(q)}\rVert_2^2},
	\end{equation}
	represents the difference between the trend values in the windows and the consensus trend value while the dual residual,
	\begin{equation*}
	r_d^{(q)} = \gamma\sqrt{\sum_{m=1}^M \sum_{j=1}^J\lVert\bar{\theta}_{j,m}^{(q)} - \bar{\theta}_{j,m}^{(q-1)}\rVert_2^2},
	\end{equation*}
	represents the change in the consensus variable from one iterate to the next. The algorithm is stopped when 
	\begin{align}
		&r_p^{(q)} < \epsilon_{abs}\sqrt{nJ} + \epsilon_{rel}\underset{m}{\max}\left[\max 
		\left(\sqrt{\sum_{j=1}^J \lVert\theta_{j,m}^{(q)}\rVert_2^2}, \sqrt{\sum_{j=1}^J \lVert \bar{\theta}_{j,m}^{(q)} \rVert_2^2} \right )\right]\\
		&r_d^{(q)} < \epsilon_{abs}\sqrt{nJ} + \epsilon_{rel}\sqrt{\sum_{m=1}^M\sum_{j=1}^J\lVert \omega_{j,m}^{(q)}\rVert_2^2}
	\end{align}

	\begin{figure}
		\centering
		\caption{Trend fit with our ADMM algorithm with 3 windows which converged in 7 iterations compared to trend from simultaneous fit.}
		\includegraphics[width = 0.8\linewidth]{Figures/admm_windows.png}
	\end{figure}

	Timing experiments illustrate the advantages of using our ADMM algorithm even on datasets where solving the problem simultaneously is possible. For each data size, $n$, 25 datasets were simulated using the peaks simulation design described below and trends for three quantiles were fit simultaneously: 0.05, 0.1, and 0.15 using a $\lambda = n/5$. We use from one to four windows for each data size with an overlap of 500. The windows algorithm was run until the stopping criteria were met using $\epsilon_{abs} = 0.01$ and $\epsilon_{rel} = 0.001$. As is shown in Fig. \ref{fig:timing}, using 4 windows instead of one on data sizes of 55000 provides a factor of 3 decrease in computation time. The timing experiments were conducted on

	\begin{figure}[!h] 
		\centering
		\caption{Timing experiments comparing quantile trend filtering with varying numbers of windows by data size.}
		\includegraphics[width = 0.7\linewidth]{Figures/Fig_timing_experiment.png}
		\label{fig:timing}
	\end{figure}

>>>>

	\if0\blind
	{
		\title{\bf Baseline Drift Estimation for Time Series Data Using Quantile Trend Filtering}
		\author{Halley Brantley\thanks{
				Department of Statistics, North Carolina State University, Raleigh, NC 27695 (E-mail: hlbrantl@ncsu.edu)} \,
			Joseph Guinness\thanks{
				Department of Statistics and Data Science, Cornell University, Ithaca, NY 14853 (E-mail: guinness@cornell.edu )} \,
			and
			%    and
			Eric C. Chi\thanks{Department of Statistics, North Carolina State University, Raleigh, NC 27695 (E-mail: eric$\_$chi@ncsu.edu).}    \\}
		\date{}
		\maketitle
	} \fi

@@

	\bigskip
	\begin{abstract}
		We address the problem of estimating smoothly varying baseline trends in time series data using quantile trend filtering. We first extend the basic framework to ensure non-crossing while estimating multiple quantile trends simultaneously. We also implement a parallelizable alternating direction method of moments (ADMM) algorithm for estimating trends in longer series. The ADMM algorthim enables the estimation of trends in a piecewise manner, both reducing the computation time and extending the limits of the method to larger data sizes. We also address smoothing parameter selection and propose a modified criterion based on the extended Bayesian Information Criterion. Through simulation studies and an application to low cost air quality sensor data, we demonstrate that our model provides better quantile trend estimates than existing methods and improves signal classification of low-cost air quality sensor output. 
	\end{abstract}

@@

	In applications spanning chemistry \citep{Ning2014}, macroeconomics \citep{yamada2017estimating}, environmental science \citep{brantley2014mobile}, and medical sciences \citep{pettersson2013algorithm, marandi2015qualitative}, scalar time series are observed and assumed to consist of a slowly varying trend and other more rapidly varying components. \cite{Kim2009} proposed \textit{$\ell_1$ trend filtering} to estimate trends that are piece-wise linear or polynomial, assuming that the series consists of a trend plus uncorrelated noise. \cite{Tib2014} later demonstrated empirically that the trend filtering estimates adapt to the local level of smoothness better than smoothing splines. In the trend filtering problem, given observations $y(t)$ with $t=1,...,N$, the trend, $\theta \in \mathbb{R}^n$, is estimated by solving the following convex problem:
	\begin{eqnarray}
	\underset{\theta}{\arg\min}\; \frac{1}{2} \lVert y - \theta \rVert_2^2 + \lambda \lVert \mathbf{D}^{(k+1)}\theta \rVert_1,
	\end{eqnarray}
	where $y \in \mathbb{R}^n$ is the vector of observations, $\lambda \geq 0$ is a regularization parameter, and the matrix $\mathbf{D}^{(k+1)} \in \mathbb{R}^{(n - k -1) \times n}$ is the discrete difference operator of order $k+1$. To understand the purpose of penalizing $\mathbf{D}^{(k+1)}$ consider the difference operator when $k = 0$.
	\begin{eqnarray}
	\mathbf{D}^{(1)} = \begin{pmatrix}
	-1 & 1 & 0 & \cdots & 0 & 0 \\
	0 & -1 & 1 & \cdots & 0 & 0 \\
	\vdots & & & & & \\
	0 & 0 & 0 & \cdots & -1 & 1 \\
	\end{pmatrix}
	\end{eqnarray}
	Thus, $\lVert \mathbf{D}^{(1)}\theta \rVert_1 = \sum_{i=1}^{n-1} \lvert \theta_{i+1} - \theta_{i} \rvert$ which is just total variation in one dimension. The penalty incentivizes solutions that are piece-wise constant. For $k \geq 1$, the difference operator $\mathbf{D}^{(k+1)} \in \mathbb{R}^{(n-k-1) \times n}$ is defined recursively as follows
	\begin{eqnarray}
	\mathbf{D}^{(k+1)} & = & \mathbf{D}^{(1)}\mathbf{D}^{(k)}.
	\end{eqnarray}
	By penalizing the $k+1$ fold composition of the discrete difference operator, we obtain solutions which are piece-wise polynomials of order $k$.

	\cite{Tib2014} showed that trend filtering is both fast and locally adaptive when the time series consists of only the trend and random noise. However, in some applications, such as the air quality monitoring problem considered in this paper, the data contain a rapidly varying signal in addition to the slowly varying trend and noise. Standard trend filtering struggles to distinguish between the slowly varying trend and the rapidly-varying signal. We propose quantile trend filtering as a more robust method for detrending in these situations (Fig. \ref{fig:trendfilter}). To estimate the trend in the $\tau$\textsuperscript{th} quantile, we minimize the criterion	
	\begin{eqnarray}
		\label{eq:quantile_trend}
		\underset{\theta}{\min}\; \rho_\tau(y - \theta) + \lambda \lVert \mathbf{D}^{(k+1)} \theta \rVert_1,
	\end{eqnarray}	
	where $\rho_\tau(z)$ is the check-loss function: 
	\begin{equation}
		\label{eq:checkloss}
		\rho_{\tau}(z) = \sum_{i=1}^n z_i(\tau-\mathbf{I}(z_i<0)),
	\end{equation}
	 and $\mathbf{I}(\cdot) $ is the indicator function. This formulation was proposed by \cite{Kim2009} as a possible extension of $\ell_1$-trend filtering but not studied. Here, we extend the basic framework to model multiple quantiles simultaneously. We also implement non-crossing constraints to ensure valid trend estimates. To reduce computation time and extend the method to larger series, we develop and a parallelizable ADMM algorithm. Finally, we propose a modified criterion for choosing the smoothing parameter.

	\begin{figure}
		\centering
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg1.png}
		\includegraphics[width = 0.45\linewidth]{Figures/trend_filter_eg2.png}
		\caption{Examples of trend filtering solutions (red) and 15\textsuperscript{th} quantile trending filtering solution (blue).}
		\label{fig:trendfilter}
	\end{figure}

	Our research is motivated by the analysis of low cost and portable air quality sensors, whose use has increased dramatically in the last decade. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. To reduce cost and power consumption of the SPod, the temperature and relative humidity of the air presented to the photoionization detectors (PIDs) is not controlled and as a result the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensor nodes co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, but the amount of baseline drift varies from one node to another.
	\begin{figure}[t!]
		\includegraphics[width = \linewidth]{Figures/uncorrected_data.png}		
		\caption{Example of 3 co-located SPod PID sensor readings.}
		\label{fig:raw_spod}
	\end{figure}

	In this application, as well as those described in \cite{Ning2014}, \cite{marandi2015qualitative}, and \cite{pettersson2013algorithm}, the goal is to estimate the trend in the baseline not the mean. We can define the trend in the baseline as the trend in a low quantile of the data. A variety of methods for estimating quantile trends have already been proposed. \cite{Koenker1978} were the first to propose substituting the squared-loss penalty with the check-loss penalty (Eq. \ref{eq:checkloss}) to estimate a conditional quantile instead of the conditional mean. Later, \cite{KoenkerNgPortnoy1994} proposed quantile trend filtering with $k = 2$ producing quantile trends that are piecewise linear, but did not extent it to higher orders. Rather than using the $\ell_1$-norm to penalize the discrete differences, \cite{nychka1995nonparametric} used the smoothing spline penalty based on the $\ell_2$-norm:

	\begin{equation*}
	\label{eq:smoothingspline}
	\rho_{\tau}(y - f) + \lambda\int (f''(t))^2 dt, 
 	\end{equation*}	
	where $f(t)$ is a smooth function of time and $\lambda$ is a tuning parameter that controls the degree of smoothing. \cite{Oh2011} proposed an algorithm for solving the quantile smoothing spline problem by approximating the check-loss function with a similar differentiable function. \cite{Racine2017} take a different approach and constrain the response to follow a smooth location scale model of the form $y(t) = a(t) + b(t)\epsilon_i$ and estimate the $\tau$\textsuperscript{th} conditional quantile using a kernel estimator and local linear approach.

	In the following sections we describe in detail our quantile trend filtering algorithms and regularization parameter criterion (Section \ref{sec:methods}). We demonstrate through simulation studies that our proposed model provides better or comparable estimates of non-parametric quantile trends than existing methods (Section \ref{sec:simluation}). We further show that quantile trend filtering is a more effective method of drift removal for low-cost air quality sensors and results in improved signal classification compared to quantile smoothing splines (Section \ref{sec:application}). Finally, we discuss potential extensions of quantile trend filtering (Section \ref{sec:discussion}).

	\section{Methods}
	\label{sec:methods}
	\subsection{Quantile Trend Filtering}

	We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$, the quantile trend filtering problem is given in Eq. \ref{eq:quantile_trend}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of methods. In many cases, including ours, we are interested in estimating multiple quantiles simultaneously. We also want to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$. Given quantiles $\{\tau_1, ..., \tau_J\}$ such that $\tau_1 < \tau_2 < ... < \tau_J$, the optimization problem becomes

	\begin{eqnarray}
	\label{eq:noncross_trend}
	\underset{\theta_1, ..., \theta_J}{\min}\; \sum_{j=1}^J \left [\rho_{\tau_j}(y - \theta_{j}) + 
	\lambda_j \lVert \mathbf{D}^{(k+1)} \theta_j \rVert_1 \right ] \\
	 \text{subject to: }\; \theta_{1}(t) \le \theta_{2}(t) \le ... \le \theta_{J}(t) \text{ for all } t,
	\end{eqnarray}	
	where $\theta_j \in \mathcal{R}^n$. The additional constraints are linear in the parameters, so the non-crossing quantile trends can still be estimated by a number of available solvers. In the rest of this paper we rely on the commercial solver Gurobi \citep{gurobi} and its R package implementation. However, we could easily substitute a free solver such as the Rglpk package by \cite{rglpk}.

@@

	The number of parameters to be estimated in this problem is equal to the number of observations multiplied by the number of quantiles of interest. As the size of the data and the number of quantiles grows, all solvers will eventually break. To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose an ADMM algorithm for solving large problems in a piece-wise fashion. The ADMM algorithm \citep{gabay1975dual, glowinski1975approximation} is fully described by \cite{boyd2011distributed}. We apply the consensus ADMM algorithm to the quantile regression trend filtering problem given in Eq. (\ref{eq:quantile_trend}), by dividing our observed series $y(t)$ with $t = \{1, ..., n\}$ into overlapping windows

	\begin{align*}
	y_1(t) = y(t) & \mbox{~~if~~} 1 \le t \le u_{1}\\
	y_2(t) = y(t) & \mbox{~~if~~} l_{2} \le t \le u_{2} \\
	\cdots & \\
	y_M(t) = y(t) & \mbox{~~if~~} l_{M} \le t \le  n\\
	\end{align*}
	 with boundaries $1 < l_{2} < u_{1} < l_{3} < u_{2} < l_{4} < u_{3} < ...< n$. An illustration is given in Fig. \ref{fig:windows}. Given quantiles $\tau_1 < ... < \tau_J$, we define $\theta_{j,m}(t)$ as the value of the $\tau_j$\textsuperscript{th} quantile trend in window $m$ at time $t$. In order to constrain the overlapping sections to be equal we define a consensus variable

	 \begin{align*}
	\bar{\theta}_{j, m}	=&  g(\theta_{j, m-1}, \theta_{j,m}, \theta_{j,m+1}) \\
	\bar{\theta}_{j, m}(t)	 =& \begin{cases} 
			 \frac{\theta_{j,m-1}(t)+\theta_{j,m}(t)}{2} & \mbox{if~~} l_{m} \le t \le u_{m-1}  \\
			 \theta_{j,m}(t) & \mbox{if~~} u_{m-1} \le t \le l_{m+1}  \\
			 \frac{\theta_{j,m}(t)+\theta_{j,m+1}(t)}{2} & \mbox{if~~} l_{m+1} \le t \le u_{m}  
			 \end{cases},
	\end{align*}
	defining $\theta_{j,M+1} = \theta_{j,M}$ and $\theta_{j,0} = \theta_{j,1}$. Our windowed quantile trend optimization problem can then be written as 
	 \begin{align*}
		 \label{eq:quantile_windows}
		 &\underset{\theta_{1,m}, ..., \theta_{J,m}}{\min}\; \sum_{m=1}^M\sum_{j=1}^J \left [\rho_{\tau_j}(y_m - \theta_{j,m}) + 
		 \lambda_j \lVert \mathbf{D}^{(k+1)} \theta_{j,m} \rVert_1 \right ] \\
		 &\text{subject to: }\; \\
		 &\qquad \theta_{1,m}(t) \le \theta_{2,m}(t) \le ... \le \theta_{J,m}(t) \text{ for all } m,t \\
		 &\qquad \theta_{j,m}(t) = \bar{\theta}_{j,m}(t) \text{ for all } j, m, t
	 \end{align*}
	 Defining Lagrange multipliers $\omega_{j,m}$, and penalty parameter $\gamma > 0$, the augmented Lagrangian for finding the trends in window $m$ is
	 \begin{equation*}
	 \mathcal{L}(\theta_{j,m}, \bar{\theta}_{j,m}, \omega_{j,m}) = \sum_{j=1}^J\rho_{\tau_j}(y_m - \theta_{j,m})+\lambda \lVert \mathbf{D}^{(k+1)}\theta_{j,m}\rVert_1 +  \omega_{j,m}^T(\theta_{j,m} - \bar{\theta}_{j,m}) + 
	 \frac{\gamma}{2}||\theta_{j,m} - \bar{\theta}_{j,m}||_2^2.
	 \end{equation*}
	 We then estimate the trend separately in each window, which can be done in parallel, while constraining the overlapping pieces of the trends to be equal as outlined in Algorithm 1. The use of the Augmented Lagrangian converts the problem from a linear program into a quadratic program. The \texttt{gurobi} R package \citep{gurobi} can solve quadratic programs in addition to linear programs, but we can also use the free R package \texttt{quadprog} \citep{quadprog}.

	\begin{figure}[!h] 
		\centering
		\includegraphics[width = 0.8\linewidth]{Figures/overlapping_windows.png}
		\caption{Window boundaries and trends fit separately in each window.}
		\label{fig:windows}
	\end{figure}

@@

	We measure convergence using the stopping criteria described by \cite{boyd2011distributed}. The criteria are based on the primal and dual residuals, which represent the residuals for primal and dual feasibility, respectively. The primal residual, 
	\begin{equation}
	r_p^{(q)} = \sqrt{\sum_{m=1}^M\sum_{j=1}^J\lVert\theta_{j,m}^{(q)} - \bar{\theta}_{j,m}^{(q)}\rVert_2^2},
	\end{equation}
	represents the difference between the trend values in the windows and the consensus trend value. The dual residual,
	\begin{equation*}
	r_d^{(q)} = \gamma\sqrt{\sum_{m=1}^M \sum_{j=1}^J\lVert\bar{\theta}_{j,m}^{(q)} - \bar{\theta}_{j,m}^{(q-1)}\rVert_2^2},
	\end{equation*}
	represents the change in the consensus variable from one iterate to the next. The algorithm is stopped when 
	\begin{align}
		&r_p^{(q)} < \epsilon_{abs}\sqrt{nJ} + \epsilon_{rel}\underset{m}{\max}\left[\max 
		\left(\sqrt{\sum_{j=1}^J \lVert\theta_{j,m}^{(q)}\rVert_2^2}, \sqrt{\sum_{j=1}^J \lVert \bar{\theta}_{j,m}^{(q)} \rVert_2^2} \right )\right]\\
		&r_d^{(q)} < \epsilon_{abs}\sqrt{nJ} + \epsilon_{rel}\sqrt{\sum_{m=1}^M\sum_{j=1}^J\lVert \omega_{j,m}^{(q)}\rVert_2^2}
	\end{align}

	\begin{figure}
		\centering
		\includegraphics[width = 0.8\linewidth]{Figures/admm_windows.png}
		\caption{Trend fit with our ADMM algorithm with 3 windows which converged in 7 iterations compared to trend from simultaneous fit.}
	\end{figure}

	Timing experiments illustrate the advantages of using our ADMM algorithm when the trend can be estimated with a single window. For each data size, $n$, 25 datasets were simulated using the peaks simulation design described below. Trends for three quantiles were fit simultaneously: 0.05, 0.1, and 0.15 using $\lambda = n/5$. We use from one to four windows for each data size with an overlap of 500. The windows algorithm was run until the stopping criteria were met using $\epsilon_{abs} = 0.01$ and $\epsilon_{rel} = 0.001$. As is shown in Fig. \ref{fig:timing}, using 4 windows instead of one on data sizes of 55000 provides a factor of 3 decrease in computation time. The timing experiments were conducted on an Intel Xeon based Linux cluster using two processor cores.

	\begin{figure}[!h] 
		\centering
		\includegraphics[width = 0.7\linewidth]{Figures/Fig_timing_experiment.png}
		\caption{Timing experiments comparing quantile trend filtering with varying numbers of windows by data size.}
		\label{fig:timing}
	\end{figure}
